Chemical tagging can work: \\ Identification of stellar phase-space structures \\ purely by chemical-abundance similarity

### 2016-01-19 papers/clusters.tex
small tweaks

% to-do
% - fix all remaining occurrences of ARC, MKN, HWR, DWH, or CITE
% - submit to arXiv
% - forward to Freeman, Bovy, others?
% - submit to journal

@@

% ARC asks: "not flagged as bad" --> what exactly do we mean here? ASPCAPFLAG = 0?

Our analysis draws on the spectra of \totalnumber\ giant stars ($\logg
< 3.9$) from \apogee\ \acronym{DR12} \citep{dr12}, all those not
flagged as bad in the \apogee\ \aspcap\ \citep{aspcap} pipeline
reductions.
We re-analyze these \emph{spectra} using \thecannon, because this
deliver stellar abundance labels of higher precision, especially for
stars of $S/N \le 150$.
The details about how we select, reduce, and analyze the \apogee\ data
are given in full detail in the companion papers \citep{casey16,
  ness16}, and we only summarize briefly here.
We re-derived 17 labels ($\teff$, $\logg$, and 15 abundances
referenced to H) form the \apogee\ \acronym{DR12} spectra. 
For the training step of \thecannon, 
12,681 red-giant stars with spectral $S/N \ge 200$ were used.
For this small fraction of the sample with the highest $S/N$ objects the
labels provided by \aspcap\ provide consistent, relatively low-scatter, sensible
abundance-space measurements.
The training step fixes a spectral model
that predicts the normalized spectrum as a quadratic function of all
labels. In the second stage---the test step---each unlabelled star is
used to establish a single-star likelihood function for its labels,
holding the model coefficients fixed.
\thecannon\ finds the labels that minimize the single-star likelihood
function for each test-set star.
This optimization is not convex, but it is embarrassingly parallel,
and therefore fast. The test
step for 150,000 spectra takes less than 30~minutes on a small 
research cluster in Cambridge.

% HWR commented this (below) out

% We start by performing our own continuum normalization on individual-exposure spectra and co-add them ourselves.
% This is required, because the performance of \thecannon\ depends
% critically on the methods used for spectral normalization.  It does
% not require accurate continuum estimation, but it requires that the
% normalization vector be obtained by a \emph{linear operation on the
%   data}. The linear operation we use is a smooth-function fit to pixels
% identified as continuum in our previous work \citep{thecannon}.

% For the training data labels (used to train \thecannon; see below), we
% use the \apogee\ \acronym{DR12} \aspcap\ atmospheric parameters and
% 15-element abundances, but only for the subset of 12,681 red-giant
% stars that have spectral signal-to-noise ratio measurements greater
% than 200.
% This is a small fraction of the total sample, but these highest
% signal-to-noise stars are observed to obtain consistent, low-scatter,
% sensible abundance-space measurements.

% As described in the original paper \citep{thecannon} and the companion
% papers \citep{casey16, ness16}, \thecannon\ is operated in two stages.
% In the first stage---the training step---the training set of stars with
% measured spectra and known labels (known parameters and abundances) is
% used to establish a likelihood function for a set of coefficients that
% parameterize a smooth, simple prediction of spectral pixels given
% labels.
% \thecannon\ finds the set of coefficients that optimizes the
% training-step likelihood function.
% In the implementation used in this \documentname, the spectral
% prediction is quadratic in labels (including all cross-terms to second
% order), and there are 17 labels ($\teff$, $\logg$, and 15 abundances
% referenced to H) and 8500 spectral pixels, so there are on the order
% of 1.4 million coefficients set by a convex optimization in the
% training step.

% In the second stage---the test step---each unlabeled star is used to
% establish a single-star likelihood function for labels, holding the
% coefficients fixed.
% \thecannon\ finds the labels that minimize the single-star likelihood
% function for each test-set star.
% This optimization is not convex, but it is embarrassingly parallel.
% For the results shown here, the training step takes hours and the test
%step for 150,000 spectra takes less than 30~minutes on a small 
%research cluster in Cambridge.

@@

Briefly, the k-means algorithm is the following:
$\bullet$~Starting at some guess for $K$ $D$-space cluster centers ($K$
locations in the $D$-dimensional space), assign each point in the
space (each star in our case) to the closest of the $K$ centers.
$\bullet$~Given that assignment of stars to centers, update each center (each of
the $K$ $D$-space positions) by taking the mean of the locations of
the stars assigned to that center.
Iterate the assignment and mean steps to convergence.
The output is the locations of the $K$ centers and the assignments of
all points to centers.
This algorithm is fast and performs well in practice in problems of
this nature; also we are not the first to use the k-means algorithm in
abundance space \citep{gratton}.

@@

Another issue with k-means is that it effectively uses metric
distances in the $D$-space; it presumes Euclidean isotropy.
We choose here to work in the hydrogen-normalized abundance space, the
space of [C/H], [N/H], [O/H], [Na/H], [Mg/H], [Al/H], [Si/H], [S/H],
[K/H], [Ca/H], [Ti/H], [V/H], [Mn/H], [Fe/H], [Ni/H].
But in addition to this, we re-scale these by approximate typical measurement
precisions obtained by \thecannon\ before running k-means.\footnote{The scalings were DWH DWH.}
This makes the space close to isotropic in measurement uncertainty or
observational precision.
Importantly we use \emph{only} abundance-space information, and no
positional or velocity information (nor $\teff$ nor $\logg$ nor any
targeting or observational meta-data) as input to the clustering
algorithm.

@@

% ARC TO DO: Make claim of the fourth being M53.
We chose a few interesting cases from the high-density clusters
in the $K=256$ run and show them in
\figurename~\ref{fig:M13}, \ref{fig:Pal5}, \ref{fig:Sgr},
\ref{fig:halo}, and \ref{fig:disk}.
The first three of these are dominated by stars in the known clusters
M13, Pal5, and Sagittarius.
The fourth is a halo structure with high velocity dispersion and
possibly accreted.
The fifth is a thin-disk star-formation feature.
We will discuss the astrophysical implications of these results in
\sectionname~\ref{sec:discussion}.

@@

\begin{figure}[!bp]
\includegraphics[width=1.5\figwidth]{./figs/clusters_0256.png}
\caption{The distribution of membership and density for the 256
  abundance-space clusters returned by the k-means algorithm at
  $K=256$.  The densest cluster (which is displayed in more detail in 
  \figurename~\ref{fig:M13}) is far more dense than the second
  densest.\label{fig:densities}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0253}
\caption{Same as \figurename~\ref{fig:all} except that the entire
  sample has been made gray and the members of the most dense
  $K=256$ abundance-space cluster have been rendered as unique, prominent triangles
  (color is velocity rank, orientation is $\logg$ rank).
  The lower-right plot shows this cluster (circle) in context of the other clusters (dots).
  This cluster, which was identified only in abundance space (six
  projections of which are the top six panels in this \figurename),
  turns out to be dominated by the halo globular cluster
  M13.\label{fig:M13}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0034}
\caption{Same as \figurename~\ref{fig:M13} but for another dense
  abundance-space cluster.
  This one turns out to be dominated by the disrupting globular
  cluster Palomar 5.\label{fig:Pal5}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0177}
\caption{Same as \figurename~\ref{fig:M13} but for another dense
  abundance-space cluster.
  This one turns out to be dominated by the Sagittarius dwarf spheroidal galaxy.\label{fig:Sgr}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0010}
\caption{Same as \figurename~\ref{fig:M13} but for another dense
  abundance-space cluster.
  This one turns out to be dominated by a high-velocity-dispersion
  structure in the Galaxy halo.\label{fig:halo}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0141}
\caption{Same as \figurename~\ref{fig:M13} but for another dense
  abundance-space cluster.
  This one turns out to be dominated by a very thin young stellar
  structure in the Galaxy disk.\label{fig:disk}}
\end{figure}

@@

The abundances of known clusters that we identify by k-means are in
excellent agreement with the literature. For the group shown in Figure \ref{fig:M13},
which we later attribute to M13, we find a mean of [Fe/H]$ = -1.55$ (DWH CHECK).
The mean and spread agree well with other work on this cluster
\citep{Kraft_1992,Cohen_Melendez_2005,Johnson_Pilachowski_2012}, and compiled catalogs of
globular cluster properties \citep[][accessed 2016]{Harris_1996}. The detailed abundances are also
consistent: we find a correlation in [C/Fe] and [N/Fe] abundances \citep{Smith_2005}, and their
projection in Figure \ref{fig:M13} show a spread that reflects deep mixing
along the red giant branch \citep{Briley_2004}. We find that the light element
abundances (most notably Mg--Al) for this cluster are anti-correlated, as
expected from other studies \citep[for example,][]{gratton}. However we find a
smaller spread in [O/Fe] than reported by others who have looked exclusively
at M13 \citep{Johnson_Pilachowski_2012}.  As we discuss above, this is expected:
k-means is by no means optimal for identifying arbitrary-shaped structures
in high-dimensional space, and the first stars that would be `discarded' from
this group would be those with the most extreme abundances: low [O/Fe] and
high [Na/Fe].

There are scant abundance measurements reported for Palomar 5, for which
most stars in the cluster in Figure \ref{fig:Pal5} belong. We find a mean
metallicity of [Fe/H]$ = -1.3$ (DWH CHECK), which lies between the only
photometric \citep{Geisler_1997} and spectroscopic \citep{Smith_1985,Smith_2002} abundance
measurements for this cluster.  The anti-correlation in C--N and Na--O abundance
have been seen elsewhere \citep[][respectively]{Smith_1985,Smith_2002}, although
to our knowledge the Mg--Al anti-correlation has not previously been shown for this cluster.
Indeed, this work contains the largest number of abundances for Palomar 5,
permitting a thorough search of tidally disrupted members \citep[for example,][]{Kuzma_2015} 
that have retained the unique cluster abundance signature.

% Sagittarius

The high-velocity-dispersion structure seen in Figure \ref{halo} stands
in stark contrast to the rest of the clusters we have identified. In
addition to being highly clustered in 15-dimensions, it is visibly
clustered in our 2-d projections. Indeed, it is more clustered in our
abundance projections than the young stellar structure in the disk,
and comparably so to the other known globular clusters that we have
shown here. The disparity between this structure and others presented
here is the lack of co-spatial stars. 12 stars within this cluster are
marked as `M53' candidates by \apogee, but the remaining 145 stars
are spread amongst fields throughout the halo.

% The high-velocity-dispersion structure in the Galaxy halo.
We freely admit that k-means is by no means optimal, and other clusters
we identify are accompanied by a few stars that are not currently co-spatial.
Those stars may be unassociated interlopers that have been misclassified,
or they may be true cluster members that are now unbound. However in the
case of the high-velocity-dispersion structure in Figure \ref{fig:halo},
the situation is far more extreme. We identify a group of stars with
highly similar abundances in 15 dimensions that are now spread throughout
the Galaxy. Because some stars are candidates of the massive globular
cluster M53, it provides tantalizing possibility that these stars may
indeed be all members of the same co-natal gas cloud, implying that 
these stars have been accreted onto the halo by a single globular
cluster early in the Milky Way's formation. While speculative (and we
make no consideration to whether this is indeed the true scenario),
this example demonstrates the ultimate goal of chemical tagging.

@@

\bibitem[Casey \etal(2016)]{ycasey16}

>>>>

% to-do
% - submit to arXiv
% - forward to Freeman, Bovy, others?
% - submit to journal

@@

Our analysis draws on the spectra of \totalnumber\ giant stars ($\logg
< 3.9$) from \apogee\ \acronym{DR12} \citep{dr12}, with no warning flags
set in the \apogee\ \aspcap\ \citep{aspcap} pipeline
reductions.
We re-analyze these \emph{spectra} using \thecannon, because this
deliver stellar abundance labels of higher precision, especially for
stars of $S/N \le 150$.
The details about how we select, reduce, and analyze the \apogee\ data
are given in full detail in the companion papers \citep{casey16,
  ness16}, and we only summarize briefly here.
We re-derived 17 labels ($\teff$, $\logg$, and 15 abundances
referenced to H) form the \apogee\ \acronym{DR12} spectra. 
For the training step of \thecannon, 
12,681 red-giant stars with spectral $S/N \ge 200$ were used.
For this small fraction of the sample with the highest $S/N$ objects the
labels provided by \aspcap\ provide consistent, relatively low-scatter, sensible
abundance-space measurements.
The training step fixes a spectral model
that predicts the normalized spectrum as a quadratic function of all
labels. In the second stage---the test step---each unlabelled star is
used to establish a single-star likelihood function for its labels,
holding the model coefficients fixed.
\thecannon\ finds the labels that minimize the single-star likelihood
function for each test-set star.
This optimization is not convex, but it is embarrassingly parallel,
and therefore fast. The test
step for 150,000 spectra takes less than 30~minutes on a small 
research cluster in Cambridge.

@@

Briefly, the k-means algorithm is the following:
$\bullet$~Start at some inital guess for the locations of $K$ $D$-space cluster centers ($K$
locations in the $D$-dimensional space).
$\bullet$~Assign each point in the
space (each star in our case) to the closest of the $K$ centers.
$\bullet$~Given this assignment of stars to centers, update each center (each of
the $K$ $D$-space positions) by taking the mean of the locations of
the stars assigned to that center.
$\bullet$~Iterate these steps (assignment of points, followed by taking of means) to convergence.
The output of this algorithm is the converged locations of the $K$ centers and the assignments of
all points to those centers.
This algorithm is fast and performs well in practice in problems of
this nature; also we are not the first to use the k-means algorithm in
abundance space \citep{gratton}.

@@

Another issue with k-means is that it effectively uses metric
distances in the $D$-space; it presumes Euclidean isotropy.
We choose here to work in the hydrogen-normalized abundance space, the
space of [C/H], [N/H], [O/H], [Na/H], [Mg/H], [Al/H], [Si/H], [S/H],
[K/H], [Ca/H], [Ti/H], [V/H], [Mn/H], [Fe/H], [Ni/H].
But in addition to this, we re-scale these by approximate typical measurement
precisions obtained by \thecannon\ before running k-means.\footnote{%
  The scalings were
  [C/H]/[0.041],
  [N/H]/[0.044],
  [O/H]/[0.037],
  [Na/H]/[0.111],
  [Mg/H]/[0.032],
  [Al/H]/[0.055],
  [Si/H]/[0.041],
  [S/H]/[0.054],
  [K/H]/[0.069],
  [Ca/H]/[0.043],
  [Ti/H]/[0.072],
  [V/H]/[0.146],
  [Mn/H]/[0.041],
  [Fe/H]/[0.019], and
  [Ni/H]/[0.034].}
This makes the space close to isotropic in measurement uncertainty or
observational precision.
Importantly we use \emph{only} abundance-space information, and no
positional or velocity information (nor $\teff$ nor $\logg$ nor any
targeting or observational meta-data) as input to the clustering
algorithm.

@@

We chose a few interesting cases from the high-density clusters
in the $K=256$ run and show them in
\figurename~\ref{fig:M13}, \ref{fig:Pal5}, \ref{fig:Sgr},
\ref{fig:halo}, and \ref{fig:disk}.
The first three of these are dominated by stars in the known clusters
M13, Pal5, and Sagittarius.
The fourth is a halo structure with high velocity dispersion and
possibly accreted.
The fifth is a thin-disk star-formation feature.
We will discuss the astrophysical implications of these results in
\sectionname~\ref{sec:discussion}.

@@

\begin{figure}[!bp]
\includegraphics[width=1.5\figwidth]{./figs/clusters_0256.png}
\caption{The distribution of membership and density for the 256
  abundance-space clusters returned by the k-means algorithm at
  $K=256$.  The densest cluster (which is displayed in more detail in 
  \figurename~\ref{fig:M13}) is far more dense than the second
  densest.\label{fig:densities}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0253}
\caption{Same as \figurename~\ref{fig:all} except that the entire
  sample has been made gray and the members of the most dense
  $K=256$ abundance-space cluster have been rendered as unique, prominent triangles
  (color is velocity rank, orientation is $\logg$ rank).
  The lower-right plot shows this cluster (circle) in context of the other clusters (dots).
  This cluster, which was identified only in abundance space (six
  projections of which are the top six panels in this \figurename),
  turns out to be dominated by the halo globular cluster
  M13.\label{fig:M13}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0034}
\caption{Same as \figurename~\ref{fig:M13} but for another dense
  abundance-space cluster.
  This one turns out to be dominated by the disrupting globular
  cluster Palomar 5.\label{fig:Pal5}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0177}
\caption{Same as \figurename~\ref{fig:M13} but for another dense
  abundance-space cluster.
  This one turns out to be dominated by the Sagittarius dwarf spheroidal galaxy.\label{fig:Sgr}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0010}
\caption{Same as \figurename~\ref{fig:M13} but for another dense
  abundance-space cluster.
  This one turns out to be dominated by a hitherto unrecognized high-velocity-dispersion
  structure in the Galaxy halo.\label{fig:halo}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0141}
\caption{Same as \figurename~\ref{fig:M13} but for another dense
  abundance-space cluster.
  This one turns out to be dominated by a very thin young stellar
  structure in the Galaxy disk, probably one that has been discovered previously \citep{wegg}.\label{fig:disk}}
\end{figure}

@@

The abundances of known clusters that we identify by k-means are in
excellent agreement with the literature. For the group shown in \figurename~\ref{fig:M13},
which we later attribute to M13, we find a mean of [Fe/H]$ = -1.55$.
The mean and spread agree well with other work on this cluster
\citep{Kraft_1992,Cohen_Melendez_2005,Johnson_Pilachowski_2012}, and compiled catalogs of
globular cluster properties \citep[][accessed 2016]{Harris_1996}. The detailed abundances are also
consistent: we find a correlation in [C/Fe] and [N/Fe] abundances \citep{Smith_2005}, and their
projection in \figurename~\ref{fig:M13} show a spread that reflects deep mixing
along the red giant branch \citep{Briley_2004}. We find that the light element
abundances (most notably Mg--Al) for this cluster are anti-correlated, as
expected from other studies \citep[for example,][]{gratton}. However we find a
smaller spread in [O/Fe] than reported by others who have looked exclusively
at M13 \citep{Johnson_Pilachowski_2012}.  As we discuss above, this is expected:
k-means is by no means optimal for identifying arbitrary-shaped structures
in high-dimensional space, and the first stars that would be `discarded' from
this group would be those with the most extreme abundances: low [O/Fe] and
high [Na/Fe].

There are scant abundance measurements reported for Palomar 5, for which
most stars in the cluster in \figurename~\ref{fig:Pal5} belong. We find a mean
metallicity of [Fe/H]$ = -1.3$, which lies between the only
photometric \citep{Geisler_1997} and spectroscopic \citep{Smith_1985,Smith_2002} abundance
measurements for this cluster.  The anti-correlation in C--N and Na--O abundance
have been seen elsewhere \citep[][respectively]{Smith_1985,Smith_2002}, although
to our knowledge the Mg--Al anti-correlation has not previously been shown for this cluster.
Indeed, this work contains the largest number of abundances for Palomar 5,
permitting a thorough search of tidally disrupted members \citep[for example,][]{Kuzma_2015} 
that have retained the unique cluster abundance signature.

The high-velocity-dispersion structure seen in \figurename~\ref{fig:halo} stands
in stark contrast to the rest of the clusters we have identified. In
addition to being highly clustered in 15-dimensions, it is visibly
clustered in our 2-d projections. Indeed, it is more clustered in our
abundance projections than the young stellar structure in the disk,
and comparably so to the other known globular clusters that we have
shown here. The disparity between this structure and others presented
here is the lack of co-spatial stars. 12 stars within this cluster are
marked as `M53' candidates by \apogee, but the remaining 145 stars
are spread amongst fields throughout the halo.

We freely admit that k-means is by no means optimal, and other clusters
we identify are accompanied by a few stars that are not currently co-spatial.
Those stars may be unassociated interlopers that have been misclassified,
or they may be true cluster members that are now unbound. However in the
case of the high-velocity-dispersion structure in \figurename~\ref{fig:halo},
the situation is far more extreme. We identify a group of stars with
highly similar abundances in 15 dimensions that are now spread throughout
the Galaxy. Because some stars are candidates of the massive globular
cluster M53, it provides tantalizing possibility that these stars may
indeed be all members of the same co-natal gas cloud, implying that 
these stars have been accreted onto the halo by a single globular
cluster early in the Milky Way's formation. While speculative (and we
make no consideration to whether this is indeed the true scenario),
this example demonstrates the ultimate goal of chemical tagging.

@@

\bibitem[Casey \etal(2016)]{casey16}
